---
title: "R Notebook"
output: html_notebook
---

```{r}
#Load libraries
library(dplyr)
library(ggplot2)
library(nycflights13)
library(resampledata3)
library(tidyverse)
```

```{r}
glimpse(flights)
```
```{r}
united_flights <- flights %>%
  filter(carrier == "UA")%>%
  filter(!is.na(dep_delay)) %>%
  filter(!is.na(arr_delay)) %>%
  filter(!is.na(air_time)) %>%
  mutate(gain = dep_delay - arr_delay)
glimpse(united_flights)
```

```{r}
late <- united_flights %>%
  filter(dep_delay >0) %>%
  filter(!is.na(dep_delay))
glimpse(late)
```

```{r}
very_late <- united_flights %>%
  filter(dep_delay >30) %>%
  filter(!is.na(dep_delay))
glimpse(very_late)
```

```{r}
united_flights <- united_flights %>%
  mutate(late = dep_delay >0) %>% 
  mutate(very_late = dep_delay >30)
glimpse(united_flights)
```

### Departure Delay Analysis(Cameron)

### Our exploratory data analysis will start with looking at a summary of arrival delay, departure delay, and net gain.

```{r}
summary(united_flights$arr_delay)
```

### Regarding arrival delay, we see that the median is -6, showing that the 50th percentile for arrival delay is showing early arrivals.
### Arrival delays show significant outliers in both directions.

```{r}
summary(united_flights$dep_delay)
```

### Regarding departure delay, we see that the median is 0, showing that the 50th percentile for departure delay cosists of flights leaving on time.
### Departure delays show significant outliers in the maximum, with a minimum that is much closer to the median and mean.

```{r}
summary(united_flights$gain)
```

### Gain gets interpreted differently than the other two facotrs here. Flights have a median of +10 minutes, meaning the 50th percentile of flights are arriving 10 minutes quicker than expexted.
### The range between the minimum and the median is quite significant, with one flight losing 2 hours and 45 minutes overall.
### There could also be some outliers at the maximum since there is a 64 minute difference between the minimum and the maximum.


```{r}
ggplot(data = united_flights, mapping = aes(x = arr_delay)) +
  geom_histogram(bins=40, color="white", fill="steelblue")
```

```{r}
ggplot(data = united_flights, mapping = aes(x = dep_delay)) +
  geom_histogram(bins=40, color="white", fill="steelblue")
```


```{r}
ggplot(data = united_flights, mapping = aes(x = gain)) +
  geom_histogram(bins=40, color="white", fill="steelblue")
```

```{r}
ggplot(data=united_flights, aes(x=late, y=gain))+
geom_boxplot(color="steelblue", fill="white")
```

```{r}
ggplot(data=united_flights, aes(x=very_late, y=gain))+
geom_boxplot(color="steelblue", fill="white")
```



### Late Hypothesis Test
### µ1 - µ2 = 0
### µ1 - µ2 != 0

```{r}
t.test(gain~late,united_flights, conf.level= 0.95)
```

### Very Late Hypothesis Test
### µ1 - µ2 = 0
### µ1 - µ2 != 0

```{r}
t.test(gain~very_late,united_flights, conf.level= 0.95)
```


