---
title: "R Notebook"
output: html_notebook
---

```{r}
#Load libraries
library(dplyr)
library(ggplot2)
library(nycflights13)
library(resampledata3)
library(tidyverse)
```

```{r}
united_flights <- flights %>%
  filter(carrier == "UA")%>%
  filter(!is.na(dep_delay)) %>%
  filter(!is.na(air_time)) %>%
  mutate(gain = dep_delay - arr_delay) %>%
  mutate(net_gain = dep_delay - arr_delay) %>%
  mutate(gain_per_hour = gain/(air_time/60)) %>%
  mutate(late = ifelse(dep_delay > 0, TRUE, FALSE)) %>%
  mutate(very_late = ifelse(dep_delay > 30, TRUE, FALSE))
united_flights

```

### Jacob (five most common destinations)

```{r}
by_destination <- united_flights %>% 
  group_by(dest) %>% 
  summarize(count = n()) %>%
  arrange(desc(count))
by_destination
```

```{r}
top_five_dest <- united_flights %>%
  filter(dest == "IAH" | dest == "ORD" | dest == "SFO" | dest == "LAX" | dest == "DEN")
```

```{r}
top_five_dest %>%
  group_by(dest) %>%
  summarize(mean = mean(net_gain)) %>%
  arrange(desc(mean))
```

```{r}
ggplot(data = top_five_dest, aes(x = dest, y = net_gain))+
  geom_point(color = "steelblue", alpha = 0.7) +
  labs(title = "Destination vs. Gain",
       x = "Destination",
       y = "Gain")
```

```{r}
ggplot(data = top_five_dest, aes(x = dest, y = net_gain)) +
  geom_col(fill = "steelblue") +
  labs(title = "Average Net Gain by Destination",
       x = "Destination Airport",
       y = "Net Gain")
```

```{r}
ggplot(data = top_five_dest, mapping = aes(x = net_gain)) +
  geom_histogram(breaks=seq(-60,60, by=5), color="white", fill="steelblue") +
  facet_wrap(~ dest)
```

**Hypothesis testing**:  Difference in average net gain between airports 

```{r}
united_iah_ord <- united_flights %>%
  filter(dest == "IAH" | dest == "ORD")
united_iah_sfo <- united_flights %>%
  filter(dest == "IAH" | dest == "SFO")
united_iah_lax <- united_flights %>% 
  filter(dest == "IAH" | dest == "LAX")
united_iah_den <- united_flights %>%
  filter(dest == "IAH" | dest == "DEN")
united_ord_sfo <- united_flights %>%
  filter(dest == "ORD" | dest == "SFO")
united_ord_lax <- united_flights %>%
  filter(dest == "ORD" | dest == "LAX")
united_ord_den <- united_flights %>%
  filter(dest == "ORD" | dest == "DEN")
united_sfo_lax <- united_flights %>%
  filter(dest == "SFO" | dest == "LAX")
united_sfo_den <- united_flights %>%
  filter(dest == "SFO" | dest == "DEN")
united_lax_den <- united_flights %>%
  filter(dest == "LAX" | dest == "DEN")
```

```{r}
t.test(net_gain~dest, data = united_iah_ord, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_iah_sfo, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_iah_lax, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_iah_den, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_ord_sfo, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_ord_lax, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_ord_den, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_sfo_lax, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_sfo_den, conf.level = .95)
```

```{r}
t.test(net_gain~dest, data = united_lax_den, conf.level = .95)
```

Clean NA 

(Jack)
Another common measure of interest, in addition to total gain, is the gain relative to the duration of the flight. Calculate the gain per hour by dividing the total gain by the duration in hours of each flight. Does the average gain per hour differ for flights that departed late versus those that did not? What about for flights that departed more than 30 minutes late?
```{r}
summary(united_flights)

```
```{r}

plot_hist(united_flights,'gain_per_hour','United Airlines flights gain per hour distribution (minutes)', bins=50)

plot_qq(united_flights,'gain_per_hour','United Airlines flights gain per hour Q-Q plot (minutes)')
```
Summary: Data strongly left skewed with majority of flights being on time.

```{r}

plot_hist(united_flights,'dep_delay','United Airlines flights departure delay distribution (minutes)',bins=50)

plot_qq(united_flights,'dep_delay','United Airlines flights departure delay Q-Q plot (minutes)')
```



```{r}

ggplot(united_flights, aes(x=dep_delay, y = gain_per_hour)) +
  geom_point(color = "cadetblue3") + 
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE, 
              color = "lightblue", 
              fill = "lightblue") 


ggplot(united_flights, aes(x=dep_delay, y = gain_per_hour, color = late)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE) +
  scale_color_manual(values = c("FALSE" = "cadetblue3", "TRUE" = "indianred"))

ggplot(united_flights, aes(x=dep_delay, y = gain_per_hour, color = very_late)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE) +
  scale_color_manual(values = c("FALSE" =  "cadetblue3", "TRUE" ="red"))




```
United Airlines flights show a clear negative relationship between departure delay and in-air time recovery (gain per hour), with a consistent negative slope across all views. On-time or early departures (red points) frequently achieve positive or near-zero gain thanks to schedule room. In contrast, very late flights (>30 min delay) rarely recover meaningful time and often lose even more relative to flight duration because possibly severe weather, airport congestion. The longer the initial delay, the less incentive or ability pilots have to “make up” time in the air. 

## Analyzing flights not very late <30min
```{r}
united_flights_delay_below_30min <-united_flights %>%
  filter(dep_delay <= 30)
```

```{r}
ggplot(united_flights_delay_below_30min, aes(x=dep_delay, y = gain_per_hour)) +
  geom_point(color = "cadetblue3") + 
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE, 
              color = "lightblue", 
              fill = "lightblue") 


ggplot(united_flights_delay_below_30min, aes(x=dep_delay, y = gain_per_hour, color = late)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE) +
  scale_color_manual(values = c("FALSE" = "cadetblue3", "TRUE" = "indianred"))


```
Once departure delay stays under ~30 minutes, United pilots neither systematically gain nor lose more. The strong negative relationship seen in the full dataset completely disappears.


## Hypothesis Analysis

Does gain per hour on flights with slightly late (<30 minutes) departure differ from flights overall?
```{r}

t.test(united_flights_delay_below_30min$gain_per_hour, mu = overall_mean, conf.level = 0.95)

```

**Result:** 
Flights that departed 30 minutes late or better recovered an average of 3.69 minutes per flight hour in the air.
United Airlines’ overall averageis 3.61 minutes per hour.
A one-sample t-test shows that flights with small or no departure delay recover slightly but significantly more time than the airline’s overall average: p = 0.022).
We are 95% confident that the true average recovery for these flights lies between 3.62 and 3.77 minutes per hour.


Do United Airlines flights that depart more than 30 minutes late recover significantly less time in the air than flights that depart within 30 minutes of schedule?
```{r}
# 4 – One-sided version of #2 (for extra credit / stronger claim)
t.test(gain_per_hour ~ very_late, data = united_flights, 
       alternative = "greater", var.equal = TRUE, conf.level = 0.95)
```
Flights delayed more than 30 minutes recover, on average, about 38 seconds per hour flown than flights delayed 30 minutes or less. A one-sided t-test confirms this disadvantage is highly significant  p ≈ 9.01e-10, and we are 95% confident the true disadvantage is at least 0.46 minutes per hour. Although there is a difference. It is less than a minute on average. A 30-minute ground delay does not translate into a dramatically worse air_time performance.



```{r}
################################################################################
#VISUALIZATION HELPERS
################################################################################

#' Quick Histogram
#' @param data Data frame or vector
#' @param var Variable name (if data is a df)
#' @param title Plot title
plot_hist <- function(data, var = NULL, title = NULL,bins = 30) {
  if(is.null(var)) {
    # data is a vector
    p <- ggplot(data = tibble(x = data), mapping = aes(x = x)) +
      geom_histogram(color = "white",bins = bins)
  } else {
    # data is a data frame
    p <- ggplot(data, aes(x = !!sym(var))) +
      geom_histogram(color = "white",bins=bins)
  }

  if(!is.null(title)) {
    p <- p + labs(title = title)
  }

  print(p)
}


#' Histogram by Group
#' @param data Data frame
#' @param var Variable name
#' @param group Group variable
#' @param title Plot title
plot_hist_grouped <- function(data, var, group, title = NULL, bins = 30) {
  p <- ggplot(data, aes(x = !!sym(var), fill = !!sym(group))) +
    geom_histogram(color = "white", position = "identity", alpha = 0.6, bins = bins)

  if(!is.null(title)) {
    p <- p + labs(title = title)
  }

  print(p)
}


#' Boxplot by Group
#' @param data Data frame
#' @param var Variable name
#' @param group Group variable
#' @param title Plot title
plot_boxplot <- function(data, var, group, title = NULL) {
  p <- ggplot(data, aes(x = !!sym(group), y = !!sym(var), 
                        fill = !!sym(group))) +
    geom_boxplot()

  if(!is.null(title)) {
    p <- p + labs(title = title)
  }

  print(p)
}


#' Q-Q Plot
#' @param data Data frame or vector
#' @param var Variable name (if data is a df)
#' @param title Plot title
plot_qq <- function(data, var = NULL, title = "Q-Q Plot") {
  if(is.null(var)) {
    # data is a vector
    p <- ggplot(data = tibble(x = data), mapping = aes(sample = x)) +
      geom_qq() + geom_qq_line()
  } else {
    # data is a data frame
    p <- ggplot(data, aes(sample = !!sym(var))) +
      geom_qq() + geom_qq_line()
  }

  p <- p + labs(title = title)
  print(p)
}


#' Bar Plot for Proportions
#' @param props Named vector or data frame with proportions
#' @param title Plot title
plot_barplot <- function(props, names = NULL, title = NULL, ylab = "Proportion") {
  if(is.null(names)) {
    names <- names(props)
  }

  barplot(props, 
          names.arg = names,
          main = title,
          ylab = ylab)
}
```


